Staged typed type-preserving tagless ``interpreters'', \emph{finally}

### 2007-04-01 tagless/tagless-final.tex
CBN CPS done

\subsection{Two tagless interpreters}
The first instantiation of the Symantics structure is an interpreter:
evaluating the object (DSL) term to its value in the meta-language. 
The following module essentially interprets (gives one-to-one
correspondence to) each operation of the object language as the
corresponding operation in the meta-language.
\begin{code}
module R  = struct
  type ('c,'dv) repr = 'dv (* absolutely no wrappers *)
  let int  (x:int)  = x
  let bool (b:bool) = b
  let add  e1 e2    = e1 + e2
  let mul  e1 e2    = e1 * e2
  let leq  x y      = x <= y
  let if_  eb et ee = if eb then (et ()) else (ee ())

@@

\jacques{Comment on online/offline here}

@@

state and imperative features

CPS transformation (interpreting/compiling the term to CBN CPS)

It's relation to PE and CPS. Currently, our CPS is CBN. Need to
investigate CBV.
The type-level function |'da -> 'sa| is important because it gives 
the CPS transform. So, the type-function
that enables the PE is exactly the same type function that enables
CPS....

The PE in incope is really a mixture of online and off-line; it looks 
online, but it has a sort of "just in time" binding-time analysis built in.

If we just add a method 'lift' to Symantics with a type
|lift :: a -> repr a| (cf polymorphic lift of Xi03, of MetaML and
MetaOCaml: CSP)
then |Symantics repr| implies |Functor repr|. We can write then
\begin{code}
  instance Semantics repr => Functor repr where
  -- fmap :: (a->b) -> repr a -> repr b
  fmap f ra = app (lift f) ra
\end{code}

>>>>

\subsection{Two tagless interpreters}
\label{S:interpreter-RL}
The first instantiation of the Symantics structure is an interpreter:
evaluating the object (DSL) term to its value in the meta-language. 
The following module essentially interprets (gives one-to-one
correspondence to) each operation of the object language as the
corresponding operation in the meta-language.
\begin{code}
module R  = struct
  type ('c,'dv) repr = 'dv (* absolutely no wrappers *)
  let int  (x:int)  = x
  let bool (b:bool) = b
  let add  e1 e2    = e1 + e2
  let mul  e1 e2    = e1 * e2
  let leq  x y      = x <= y
  let if_  eb et ee = if eb then (et ()) else (ee ())

@@

\jacques{Comment on online/offline here}
The PE in incope is really a mixture of online and off-line; it looks 
online, but it has a sort of "just in time" binding-time analysis built in.

@@

\oleg{where to put the following? Should we even mention this?
If not, we should still keep this in comments.}
If we just add a method 'lift' to Symantics with a type
|lift :: a -> repr a| (cf polymorphic lift of Xi03, of MetaML and
MetaOCaml: CSP)
then |Symantics repr| implies |Functor repr|. We can write then
\begin{code}
  instance Semantics repr => Functor repr where
  -- fmap :: (a->b) -> repr a -> repr b
  fmap f ra = app (lift f) ra
\end{code}

\subsection{Call-by-name CPS interpreters}

The object language generally inherits the evaluation strategy from
the metalanguage. Evaluating an application |e1 e2| in OCaml requires
the evaluation of |e2| before performing the application. Therefore,
when evaluating the encoding of the object language application 
|app e1 e2|, |e2| will always have to be evaluated. One may wonder if it is
possible then to represent a call-by-name (CBN) object language in a
call-by-value (CBV) metalanguage. The answer, given back by Reynolds
and Plotkin (cite) is CPS, which was specifically introduced by them
to make the evaluation strategy of the definitional interpreter
independent of the strategy of the metalanguage. We demonstrate the
same independence in the typed setting \oleg{mention the quote from
  Reynolds that typed setting would be nice?}. Specifically, we
demonstrate a CBN evaluation of our object language in OCaml, which is
CBV. We do so by building a CBN CPS interpreter for the object
language.

The CBN CPS interpretation was introduced by Plotkin (cite). As in the
more familiar CBV CPS, the interpretation of a value is a function
accepting the continuation argument |k| and yielding the answer
returned by that continuation.
\begin{code}
let int (x:int) = fun k -> k x
let add e1 e2 = 
  fun k -> e1 (fun v1 -> e2 (fun v2 -> k (v1+v2)))
\end{code}
In the both cases, the returned value has the type 
|(int -> 'w)->'w| where |'w| is the (polymorphic) answer type.

The interpretation of abstraction and application in CBN CPS is
different from CBV CPS and is as follows:
\begin{code}
let lam f = fun k -> k f
let app e1 e2 = fun k -> e1 (fun f -> (f e2) k)
\end{code}
In the tell-tale sign of CBN, in interpreting the application
we do not `evaluate' the argument e2, that is, we do apply it to the
continuation. Rather, we pass the unevaluated |e2| to the abstraction.
The result of |lam (fun x -> add x (int 1))| has the type
|((((int->'w1)->'w1) -> ((int->'w1)->'w1)) -> 'w2) -> 'w2|.
We want to collect the above interpretation functions into a module
with the Symantics signature, so to fit the CBN CPS interpreter within our
general framework. Alas, we encounter the same problem we had in
Section~\ref{S:PE-problem}, in writing a type equation for the |repr|
type that can represent both the type of |int i| and the type of 
|lam f|. We cannot say that the the result of interpreting an object
expression of the type |t| is |(t->'w)->'w| because if |t| happens to
be an arrow type such as |int->int| in the above example, the result
type obviously does not fit the desired pattern. We need a type
function with a type-case distinction. In Section~\ref{S:PE-solution} we have
shown the way to encode such a function, using the additional type
argument |'sv| of the |repr| type. Furthermore, we notice that the type 
function |repr_pe| needed for the partial evaluator,
Section~\ref{S:PE-solution}, is precisely the same type function we
need for CBN CPS. This gives us for the following CBN CPS interpreter,
of the Symantics signature:

\begin{code}
module RCN = struct
  type ('c,'sv,'dv) repr = 
     {ko: 'w. ('sv -> 'w) -> 'w}
  let int (x:int) = {ko = fun k -> k x}
  let add e1 e2 = 
    {ko = fun k -> e1.ko (fun v1 -> 
                   e2.ko (fun v2 -> k (v1+v2)))}
  let if_ eb et ee = 
    {ko = fun k -> eb.ko 
         (fun vb -> if vb then (et ()).ko k 
                    else (ee ()).ko k)}
  ...
  let lam f = {ko = fun k -> k f}
  let app e1 e2 = 
     {ko = fun k -> e1.ko (fun f -> (f e2).ko k)}
  let fix f = 
     let rec fx f n = app (f (lam (fx f))) n in 
     lam(fx f)

  let get_res x = x.ko (fun v -> v)
end
\end{code}

We have made our interpreter fully polymorphic over the answer type.
We could have just as well made |RCN| a functor, parameterized over
the answer type -- as is common when writing CPS in SML. In the
present case, we chose to use higher-rank polymorphism of OCaml record
types.

Because |RCN| has the Semantics signature, we can instnatiate with it and run
all of our previous examples. A more interesting is the following
example.
\begin{code}
module EXS(S: Symantics) = struct
 open S
 let diverg () = 
   app (lam (fun x -> int 1)) 
       (app (fix (fun self -> self)) (int 2))
end
\end{code}

If we interpret |EXS| with the |R| interpreter of
Section~\ref{S:interpreter-RL} realized in OCaml
\begin{code}
let module M = EXS(R) in M.diverg ();;
\end{code}
we never get any result because of looping. In contrast, if we use
the CBN interpreter
\begin{code}
let module M = EXS(RCN) in RCN.get_res (M.diverg ())
\end{code}
we get the result |1|.

\oleg{mention some formal results? Like indifference theorems by
  Plotkin? And say we generalized their result to the type setting?
  Mention this in the conclusions?}

\subsection{CBV CPS transformers}

We can obtain CBV from CBN?

\subsection{State and imperative features}
